First paper - Draft

### 2017-06-21 paper/paper.Rnw
Wrote research design section + plausible values

@@

But the most importat variable for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary). This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Talk about plausible values, what they mean and how they're measured.

@@

@@

This implementation has several drawbacks (talk about the assumptions that reardong makes).

We estimate the SES gap in a very simple fashion. For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile. Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate. We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.

@@

<<variance_pisa, cache = TRUE>>=

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer() %>%
  ggplot(aes(name, value)) + geom_col()
@
\subsection{Results}

>>>>

This paper is interested in two questions. As explained in the literature review, we already know that U.S achievement gaps have been growing over time, and some international evidence suggests that it might be increasing in other countries as well. All of these studies pool all students together from different ages and estimate global trends.

We take a different approach and we want to study age-specific trends in achievement gaps. We're interested in studying a) the size of the achievement in a comparative perspective and its relationship to overall achievement levels, b) which countries are experiencing changes in the achievement gap, c) the rate at which each country-gap is widening/narrowing and d) establishing whether the gaps are widening/narrowing because particular groups are getting ahead/behind.

We develop each question separately for more detail.

\begin{description}
  \item[We want to test the notion that better performing countries have lower leves of achievement gap. \citep{werfhorst_mijs} emphasize that there is empirical evidence that suggests this. This pattern is not so obvious. For example, countries with high leves of tracking could maximize student performance, specially the high SES students, raising their overall performance and thus raising the national performance score. But if the bottom performers are not gaining at the same rate, then the achievement gap will inevitable grow resulting in a high performing countries with widening achievement gap.]
  \item[The seminal work of \citet{reardon2011} suggests that achievement gaps change, and they do so much quicker that we though. After recording a SES gap increase of about 40\% in only 30 years, \citep{reardon_portilla} stress that they also found a significant decrease in only 15 years of data. This lends credibility that our estimations will show variation over time. As usual, their estimations are only for children going deom 4th to 8th grade. We concentrate on age-specific groups, in this case, 15 year olds.]
  \item[We want to compare the percentage change at which the gap widened/narrowed from the first to the last year. This will give us a general idea of the overall change over time, and will allow us to compare our estimates to the actual literature \footnote{Although no study has performed this age-specific achievement gap for comparable tests over such a long time. Our results will serve as comparison for other studies that use age-specific groups, such as 4th graders.}]
  \item[The widening/narrowing of the achievement gap has a source, which has been often studied to be related to everything from educational spending, income inequality, time allocation to students and preschool enrollment. The literature has concentrated very narrowly on whether the gap is increasing because the top performers are getting ahead, because both are distancing or because the bottom is falling behind. We shall pay particular attention to identifying the rate at which the top/bottom groups are evolving over time. This type of analysis is particularly useful in jotting down the mechanisms through which the gap is evolving. If the bottom performers are stagnated whereas the top is gaining ground, the efficiency-trade off dilemma of tracking might be much more credible than other explanations.]
\end{description}

@@

But one of the most relevant variables for our study is the is a composite SES index created by the PISA team. The index of economic, social and cultural status (ESCS) was created on the basis of the following variables: the International Socio-Economic Index of Occupational Status (ISEI); the highest level of education of the student’s parents, converted into years of schooling; the PISA index of family wealth; the PISA index of home educational resources; and the PISA index of possessions related to "classical" culture in the family home (Education at a Glance, OECD, Paris, 2002, Glossary). This variable, aside from capturing all relevant dimensions of SES, such as education, occupation, and material resources, takes care of transforming the variables into comparable measures across waves. In fact, we use the newly-released ESCS index (% http://www.oecd-ilibrary.org/education/pisa-2015-results-volume-i/pisa-2015-technical-background_9789264266490-13-en
) which rescaled all ESCS indexes to make them suitable for over-time analysis \footnote{These rescaled indices can be found in the \href{http://www.oecd.org/pisa/data/2015database/}{PISA website} under \emph{Rescaled Indices for Trend Analyses}.}.

Aside from SES, the other most relevant variables are test scores. PISA provides plausible values instead of one single indicator of test score performance. As explained in the PISA manual (PISA 2012, 2015), these are imputed values that resemble individual test scores and have approximately the same distribution as the latent trait being measured. Suppose we have µi, the average student test score in mathematics for student i. Instead of estimating µi alone, plausible values estimate a distribution of possible µ’s, for student i, together with the likelihood of each µi based on the respondents answers on the test. This is defined as the posterior distributions of µ’s for student i. The reason why we use this procedure is because estimating a single estimate µi is plagued with measurement error, among other types of bias (see Wu, 2005). The plausible values in the PIAAC data are effectively five (ten for PISA 2015) random draws from this distribution. These variable is are continuous, ranging from 0 to 500, with a mean of 250.

@@

% Transform all of this text into equations

@@

Standardizing the median test score solves the problem of comparability of gaps measured with different tests, and across waves. However, if the variance of academic achievement changes over time, then standardizing the overall score at each wave of testing confounds changes in the gap with changes in the variance of test scores. That is, by standardizing we're forcing the variance of test scores to be zero across all waves. But if the true variance of academic achievement grows over time, then the estimated trend in the achievement gaps will be underestimated, and vice versa.

% Show graph of variance over time
% I'm showing standard deviations, is it better to show variance?
<<variance_pisa, cache = TRUE>>=

var_years %>%
  setNames(seq(2000, 2015, 3)) %>%
  enframer() %>%
  ggplot(aes(name, value)) + geom_col()
@

Inspection of the data suggests that it's something we shouldn't be deeply concerned with. The standard deviation of each wave seems to be following a very similar pattern with an exception of the year 2000.

Another concern is that if the gaps at different waves are measured with tests that have different amounts of measurement error, then the amount of bias will not be the same in each measure of the gap. This can be very misleading and suggest erroneous intepretations regarding trends in the of the gaps over time \citet{reardon2011}. The PISA has tried to make sure the tests are comparable across waves (PISA 2012 technical report), but still we have to adjust for this imprecision.

In order to correct gap estimates for measurement error, I weight the student test scores by the P where P is the reliability of the test for each wave. Each PISA study provided a global reliability indicator which we use accordingly to each wave. This yields estimates of the true gaps, and eliminates any bias in the trend that may arise from differential reliability of the tests.

After constructing the adjusted test score measurement, we estimate the SES gap in a very simple fashion. For each country-wave, we estimate the weighted 90th and 10th percentile scores for the SES composite index, and then we generate a dummy of 1 for those above (including) the 90th percentile and 0 for those below (including) the 10th percentile. Finally, for each wave, we fit a mixed model with the adjusted test score as the dependent variable and the SES index as the only covariate. We allow both the intercept and the dummy variable to vary by country, and weight the model by the appropriate student weights.

@@

\subsection{Results}
